Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-08 Rebuttal.md

# Rebuttal

## Reviewer 1.
- Applicability: Comparison to standard workflow is not possible. The classic epidemiological workflow is strictly hypothesis driven and therefore requires a small set of 2 - 3 variables, which are correlated with a target variable. IVA methods amplify *hypothesis generation* and therefore aim to provide insight into unexpected relationships in the data.

### 2014-12-08 draft.tex
revised results figure
revised Summary and Conclusion section

Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.

@@

Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.

@@

\begin{figure*}[p!]

@@

  \includegraphics[width=0.95\textwidth]{figures/results}

@@

no group (a), subdivision into \emph{male} and \emph{female} subjects (b), quartiles of \emph{Body Mass Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).

@@

We also create subject groups to assess the influence of variables affecting the lumbar spine shape.

@@

We carry out the analysis using different subject groups:
\begin{enumerate}
	\item All subjects,
	\item subdivision into \emph{male} and \emph{female},
	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
\end{enumerate}

@@

The first plot shows all original variables.

@@

The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).

@@

\subsection{All Variables}

@@

The vast majority of non-image variables cannot be automatically classified based on the image-variables.

@@

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
%%
The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

The analysis of the dummy variables yields a result similar to the \emph{blood pressure medication}.
%%
Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields an error below $0.5$.
%%
\subsection{Gender Groups}
%%
Classification using groups divided by \emph{gender} do not produce satisfying results.
%%
Only \emph{hypothyroidism} could be described for male subjects with an error of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
%%
\subsection{Body Mass Index Groups}
%%
\emph{Gender} could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.
%%
The result however, is probably overfitted to the data due to tree sizes between $14$ and $16$.
%%

Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to \emph{back} and \emph{leg pain} due to a more stressed lumbar spine.
%%
The stress-induced spine deformation seems to directly influence the pain levels for these subjects.
%%
The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
%%
\\\\
\noindent \textbf{Large Decision Trees.}
Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\\\\
\noindent \textbf{Small Decision Trees.}
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
%%
%TODO: [medium] Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
%%
\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
%%
They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
%%
\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.

Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
%%
The remaining information within these groups is not sufficient to characterize back pain-related variables.
%%
Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
%%

% Another observation during the analysis is that large decision trees achieve low error scores.
%%
% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
%%
%\\\\
\noindent \textbf{Applicability.}
%%
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
%%
Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our interactive visualization.
%%
Therefore, the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
%%
\\\\
\noindent \textbf{Predictive Power of Image-Derived Variables.}
We showed that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
%%
Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
%%
Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
%%
\\\\
\noindent \textbf{Future Work.}
%% Contributions
We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%% TODO: [high] This is no related work
We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%
The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.

The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
%
These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.

%% TODO: [high] Here starts the future work
In our future work, we will focus on more precise models for extracting measures.
%%
Dented vertebras are an early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
%%
Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
%%
As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
%%
\\\\
%% TODO: [high] This is a summary, not future work
Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
%%
These methods do not aim to replace the traditional epidemiological workflow, but rather intervene at the weak points of standard statistical methods.
%%
Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.

\section*{\uppercase{Acknowledgements}}
\begin{small}
\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
\end{small}

>>>>

%Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.

@@

Further subdivision, e.g. by quantiles of \emph{body mass index} increases the number to 402 trees.

@@

% \begin{figure*}[p!]
\begin{figure*}[!t]

@@

  \includegraphics[width=1\textwidth]{figures/results}

@@

no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c). % and quartiles of \emph{body size} (d).

@@

We create subject groups to assess the influence of variables affecting the lumbar spine shape.

@@

The groups are: all subjects; \emph{males} and \emph{females}; and a subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48].
% We carry out the analysis using different subject groups:
% \begin{enumerate}
% 	\item All subjects,
% 	\item subdivision into \emph{male} and \emph{female},
% 	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
% 	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
% \end{enumerate}

@@

The first plot shows all original variables, the second all categorical variables transformed into dichotomous \emph{dummy variables}.

@@

%The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).

@@

\noindent \textbf{All Variables.}

@@

The majority of non-image variables cannot be automatically classified based on the image-variables.

@@

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all medicated subjects were correctly classified ($796$ of $1,058$).
%%
The majority of non-medicated subjects are false-positive classified ($262$ of $1,058$), yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

The analysis of the dummy variables yields a result similar to the \emph{blood pressure medication}.
%%
Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields an error below $0.5$.
%%
\\\\
\noindent \textbf{Gender Groups.}
%%
Classification using groups divided by \emph{gender} do not produce satisfying results.
%%
Only \emph{hypothyroidism} could be described for male subjects with an error of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
%%
\\\\
\noindent \textbf{Body Mass Index Groups.}
%%
\emph{Gender} could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
%%
The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
%%
The starting age of smoking could be characterized well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.
%%
The result however, is probably overfitted to the data due to tree sizes between $14$ and $16$.
%%

Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.
%%
Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
%%
The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
%%
The resulting tree consists of $8$ rules and has an error of $0.35$.
%%
Obese subjects are prone to \emph{back} and \emph{leg pain} due to a more stressed lumbar spine.
%%
The stress-induced spine deformation seems to directly influence the pain levels for these subjects.
%%
The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
%%
% \subsection{Size Groups}
% %%
% Many previously described results are influenced by \emph{subject size}.
% %%
% Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
% %%
% For example, large subjects are already characterized by their rather straight spine.
% %
% The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
% %%
% Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
% %%
% \\\\
% \noindent \textbf{Large Decision Trees.}
% Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
% %%
% \emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
% %%
% For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
% %%
% There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
% %%
% Similar values can be observed for \emph{cholesterol} or \emph{age}.
% %%
% Due to the large decision trees, these results are not usable and impose overfitting to the data.
% %%
% \\\\
% \noindent \textbf{Small Decision Trees.}
% %%
% The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
% %%
% Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
% %%
% These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
% %%
\section{\uppercase{Summary \& Conclusion}}
\label{sec:Conclusion}
%% Contributions
\noindent We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%% TODO: [high] This is no related work
We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%
The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.

The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
%
These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.
% \noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
% They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
% This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
% \emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
% Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
% The remaining information within these groups is not sufficient to characterize back pain-related variables.
% Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
% Another observation during the analysis is that large decision trees achieve low error scores.
% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
\noindent \textbf{Predictive Power of Image-Derived Variables.}
The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
\noindent \textbf{Applicability.}
Methods supporting \emph{hypothesis generation} based on image information are new to the application domain.
They are an addition to the standard epidemiological workflow as they highlight new possibly unknown relationships.
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
Huge result spaces could be navigated fast using our interactive visualization.
Therefore, the method is applicable not only for deriving information based on image data, but on all \emph{potential target variables}.
% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
% Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
% This can be achieved by creating a describe several (dichotomous) target values with a decision tree constructed from all available data
% We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
% This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
\noindent \textbf{Future Work.}
We will focus on more precise models for extracting measures.
Dented vertebras hint a pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
Spine canal thickness indicates signs of herniated disc disease and is also of interest.
%Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
We aim to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
\noindent \textbf{Outlook.}
Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
These methods do not aim to replace the traditional epidemiological workflow, but rather complement the weak points of standard statistical methods.
Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.
\section*{\uppercase{Acknowledgements}}
\begin{small}
\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
\end{small}
